\section{Discussion and Limitations}
\label{sec:discussion}

\paragraph{What IsoColBERT does and does not do.} The retrieval gains concentrate where two conditions intersect in our setup: weak pretrained signal and retrieval headroom. High-resource pairs sit at the P@1 ceiling for both methods, and the small ($-0.05$ pp on EN-ES, well within the seed-wise standard deviation) shift on the training pair shows that the regularizer does not damage in-distribution accuracy on this benchmark. The regularizer reshapes geometry in proportion to its weight $\lambda$, but the retrieval payoff saturates within our $\lambda \in \{0, 0.1, 0.5\}$ grid once anisotropy stops being the bottleneck. We do not have evidence on what would happen at much larger $\lambda$, on different benchmarks, or on different backbones.

\paragraph{Asymmetric collapse under bitext contrastive training.} An incidental finding that we have not seen reported elsewhere: under symmetric bitext InfoNCE, the target-side language collapses harder than the source-side language. Spanish tokens have intra-cosine $0.29$ under vanilla training while English tokens have $0.20$; effective ranks are $101.7$ and $105.1$ respectively. We attribute this to the geometry of pretraining: XLM-R is exposed to far more English than Spanish text, so the contrastive objective has more work to do on the Spanish side and the corresponding representations bear most of the gradient pressure. IsoColBERT closes the gap. The geometry of training-pair language asymmetry suggests a diagnostic for predicting which side of a contrastive setup will benefit most from explicit isotropy pressure.

\paragraph{Relation to I-STAR.} \citet{rudman2024istar} report that for classification fine-tuning, decreasing isotropy can improve accuracy; our result on retrieval is in the opposite direction. We do not run a head-to-head experiment, so we cannot identify which specific factor accounts for the discrepancy. Two candidate explanations are: (1) classification benefits from local clustering (same-class examples should be nearby, which is locally anisotropic), whereas late-interaction retrieval benefits from angular separation across the full manifold so that the argmax in MaxSim has high cosine margin; and (2) classification operates on sentence-pooled representations while late-interaction retrieval operates on individual token vectors, so the geometric quantity that matters is different. These are plausible but not tested. The narrower claim we can defend is: the productive direction for isotropy regularization appears to be task-dependent, and in our cross-lingual late-interaction setting it is toward higher isotropy.

\paragraph{Scope and generalization.} All claims in this paper are evaluated on a single benchmark family (FLORES+) and a single backbone (XLM-RoBERTa-base), trained from a single source pair (EN-ES). We have not tested whether the effect transfers to: (i) harder retrieval benchmarks such as MIRACL \citep{zhang2023miracl} or mMARCO \citep{bonifacio2021mmarco}, which use real user queries and adversarial distractors rather than parallel news translations; (ii) larger multilingual encoders or retrieval-specialized backbones such as multilingual E5 \citep{wang2024me5}, which may have less initial anisotropy and therefore less headroom for the regularizer; (iii) multi-pair or curriculum training, which could change which side bears most of the contrastive pressure; (iv) downstream tasks beyond MaxSim retrieval. Until those tests exist, we cannot claim that the geometric pathology we measure is the dominant retrieval bottleneck in general, nor that the regularizer would help under different evaluation conditions.

\paragraph{Methodological limitations.} Three seeds is the practical minimum for reporting variance; we note per-seed values throughout to allow audit. The $\lambda$ grid is coarse ($\{0, 0.1, 0.5\}$); finer-grained search around $\lambda=0.5$ may yield additional retrieval payoff. We present mechanism ablations that compare three regularizer placements but did not test other natural placements (regularizing only document tokens, restricting to a percentile of tokens, or layer-specific regularization). Our effect-size markers (\textbf{***/**/*}) are cutoffs against the seed-wise standard deviation and should not be read as statistical significance: with three seeds a paired test has only two degrees of freedom, too few for conventional thresholds, and a per-query bootstrap or randomization test would be a more principled replacement that we leave to future work.
